### 2016-04-25 text/Chapter04.tex
Finish edits (Chapter 4)

\noindent
In two of the three systems, some of the operators of the flat coeffect algebra coincide, but
in the dataflow system all three are distinct. Similarly, the two special elements
coincide in some, but not all systems. The required laws are motivated by the aim to capture
common properties of the three examples, without unnecessarily restricting the system:

\begin{itemize}
\item The monoid $(\C, \cseq, \cunit)$ represents \emph{sequential} composition of (semantic)
functions. The laws of a monoid are required in order to form a category structure in the
semantics (Section~\ref{sec:flat-semantics}).

\item The idempotent monoid $(\C, \cpar, \czero)$ represents \emph{pointwise} composition,
\ie~the case when the same context is passed to multiple (independent) computations. The monoid
laws guarantee that usual syntactic transformations on tuples and the unit value
(Section~\ref{sec:flat-exts}) preserve the coeffect. Idempotence holds for all our examples
and allows us to unify the flat and structural systems in Chapter~\ref{ch:unified}.

@@

\noindent
For simplicity, assume that all parameters have the same type $\ident{num}$ and so the annotations only
track sets of names. The definition uses a set union for all three operations. Both variables and
constants are are annotated with $\emptyset$ and the ordering is defined by $\subseteq$. The
definition satisfies the flat coeffect algebra laws because $(S, \cup, \emptyset)$ is an idempotent,
commutative monoid. The language has additional syntax for defining an implicit parameter and for
accessing it, together with associated typing rules:
%
\begin{equation*}
\quad~ e ~::=~ \ldots \sep \ident{?p} \sep \kvd{let}~\ident{?p}~=~e_1~\kvd{in}~e_2
\end{equation*}
\begin{equation*}
\tyrule{param}
  { ~ }
  { \coctx{\Gamma}{\cclrd{ \{ \ident{?p} \} }} \vdash \ident{?p} : \ident{num} }
\end{equation*}
\begin{equation*}
\tyrule{letpar}
  { \coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 &
    \coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_2}
  {\coctx{\Gamma}{\cclrd{r} \;\cclrd{\cup\; (\cclrd{s} \setminus \{ \ident{?p}) \}} } \vdash \kvd{let}~\ident{?p}=e_1~\kvd{in}~e_2 : \tau_2 }
\end{equation*}
%
The (\emph{param}) rule specifies that the accessed parameter $\ident{?p}$ needs to be in the set of
required parameters $\cclrd{r}$. As discussed earlier, we use the same type $\ident{num}$ for all
parameters, but it is also possible to define a coeffect calculus that uses mappings from names to
types (care is needed to avoid assigning multiple types to a parameter of the same type).

@@

As required by the laws, $(\mathbb{N}, +, 0)$ and $(\mathbb{N}, \mathit{max}, 0)$ form monoids
and $(\mathbb{N}, \mathit{min})$ forms a band. Note that dataflow is our first example where
$\cseq$ is not idempotent. The distributivity laws require the following to be the case:
$\mathit{max}(r,s) + t = \mathit{max}(r+t, s+t)$, which is easy to see.

@@

\noindent
Note that $(\mathbb{N}_{\bot}, +, 0)$ is a monoid for the extended definition of $+$; for the
bottom element $0 + \bot = \bot$ and for natural numbers $0 + n = n$. The structure
$(\mathbb{N}, \emph{max}, \bot)$ is also a monoid, because $\bot$ is the least element and so
$\mathit{max}(n, \bot) = n$. Finally,  $(\mathbb{N}, \emph{min})$ is a band (the extended
\emph{min} is still idempotent and associative) and the distributivity law also holds
for $\mathbb{N}_{\bot}$.

@@

The flat coeffect algebra introduced in Section~\ref{sec:flat-calculus} requires a number of laws.
The laws are required for three reasons -- to be able to define the categorical structure
in Section~\ref{sec:semantics-theory}, to prove equational properties in Section~\ref{sec:flat-syntax}
and finally, to guarantee intuitive syntactic properties for constructs such as
$\lambda$-abstraction and pairs in context-aware calculi.

@@

Recall the (\emph{let}) rule in Figure~\ref{fig:flat-types}. It annotates the expression
$\kvd{let}~x=e_1~\kvd{in}~e_2$ with context demands $\cclrd{s}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{r})$.
This rule can be derived from the typing derivation for an expression $(\lambda x.e_2)~e_1$ as
a special case, by using the idempotence of $\czip$ as follows:
%
\begin{equation*}
\tyrule{app}
  {\begin{array}{l}
   \vspace{-1.5em}
   \coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1
   \end{array} &
   \tyruler{abs}
       { \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{s}} \vdash e_2 : \tau_2 }
       { \coctx{\Gamma}{\cclrd{s}} \vdash \lambda x.e_2 : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 } }
  { \coctx{\Gamma}{\cclrd{s}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{r})} \vdash (\lambda x.e_2)~e_1 : \tau_2 }
\end{equation*}
%
This design decision is motivated by the fact that the typing obtained using the special rule for
let-binding is more precise for all the examples consider in this chapter. To see this, assume
an arbitrary splitting $\cclrd{s} = \cclrd{s_1}\,\czip\,\cclrd{s_2}$. Table~\ref{tab:flat-simplelet}
shows the coeffect annotation derived from $(\lambda x.e_2)~e_1$, the coeffect annotation obtained
by the (\emph{let}) rule and the simplified coeffect annotation using the particular flat coeffect
algebras.

@@

It is perhaps somewhat unexpected that the annotation can be simplified in different
ways for different examples. However, for all our systems, the simplfied annotation
(on the right) is more precise than the original (on the left). Recall that
$\cclrd{s} = \cclrd{s_1}\,\czip\,\cclrd{s_2}$. The following three inequalities hold:
%
\begin{equation*}
\begin{array}{rclll}
\cclrd{s_1} \cup (\cclrd{s_2} \cup \cclrd{r}) &\supseteq& (\cclrd{s_1} \cup \cclrd{s_2}) \cup \cclrd{r}
  && \textnormal{(implicit parameters)}\\
\cclrd{s_1} \sqcap (\cclrd{s_2} \sqcup \cclrd{r}) &\sqsupseteq&  (\cclrd{s_1} \sqcap \cclrd{s_2})
  && \textnormal{(liveness)} \\
\mathit{max}(\cclrd{s_1}, \cclrd{s_2} + \cclrd{r}) &\geq& \mathit{min}(\cclrd{s_1}, \cclrd{s_2}) + \cclrd{r}
  &\quad& \textnormal{(dataflow)} \\
\end{array}
\end{equation*}
%
In other words, the inequality states that using idempotence, we get a more precise typing.
Using the $\cgeq$ operator this property can be expressed using the abstract operators of the flat coeffect algebra as:
%
\begin{equation*}
\cclrd{s_1} \;\cpar\; (\cclrd{s_2} \,\cseq\, \cclrd{r}) \;\cgeq\;
  (\cclrd{s_1} \,\czip\, \cclrd{s_2}) \;\cpar\; ((\cclrd{s_1} \,\czip\, \cclrd{s_2}) \,\cseq\, \cclrd{r})
\end{equation*}
%
This property does not follow from the axioms of the flat coeffect algebra. To make
the flat coeffect system as general as possible, we do not \emph{in general} require it as
an additional axiom, although the above examples provide a reasonable basis for using
the specialized (\emph{let}) rule in the flat coeffect system.

@@

In Section~\ref{sec:flat-calculus-lambda}, we discussed how to reconcile two typings for
lambda abstraction -- for implicit parameters, the lambda function needs to split context
demands using $\cclrd{r} \cup \cclrd{s}$, but for dataflow and liveness it suffices to
duplicate the demand $\cclrd{r}$ of the body. We introduced the $\czip$ operation as
a way of providing the additional abstraction.

In this section, we first identify coeffect calculi for which the simpler (\emph{idabs}) rule
introduced in Section~\ref{ec:flat-unique-ldf} is sufficient. Then we look at syntactic
transformations corresponding to other common properties of $\czip$.

\paragraph{Simplified abstraction.}
Recall that $(\C, \czip)$ is a band, that is, $\czip$  is idempotent and associative. The
idempotence means that the context demands of the body can be required from both the
declaration site and the call site. Thus, the (\emph{idabs}) typing is valid. For reference,
we repeat the earlier definitions here:

@@

\noindent
The practical consequence of the Remark~\ref{thm:flat-alt-abs} is that, for strictly
oriented coeffect calculi (such as our liveness and dataflow computations), we can use
the (\emph{idabs}) rule and get an equivalent type system. This alternative formulation
removes the non-determinism of type checking that arises from the splitting of context
demands in the original (\emph{abs}) rule.

@@

\noindent
Intuitively, this means that the context demands of a function with multiple arguments can be
split arbitrarily between the declaration site and (multiple) call sites. In other words, it does
not matter how the context demands are satisfied.

@@

To show the key aspects of flat coeffect systems, the calculus introduced in Section~\ref{sec:flat-calculus}
consists only of variables, abstraction, application and let binding. Here, we extend it
with pairs and the unit value to sketch how it can be turned into a more complete programming
language and to motivate the laws required about $\cpar$. The syntax of the language is extended as follows:

\noindent
\begin{equation*}
\begin{array}{rcl}
e &::=& \ldots \sep () \sep e_1, e_2 \\
\tau &::=& \ldots \sep \ident{unit} \sep \tau \times \tau
\end{array}
\end{equation*}
%
The typing rules for pairs and the unit value are shown in Figure~\ref{fig:flat-ext-types}.
The unit value (\emph{unit}) is annotated with the $\czero$ coeffect (the same as other constants).
Pairs, created using the $(e_1, e_2)$ expression, are annotated with a coeffect that combines
the coeffects of the two sub-expressions using the \emph{pointwise} operator $\cpar$. The operator
models the case when the (same) available context is split and passed to two independent
sub-expressions. Finally, the (\emph{proj}) rule is uninteresting, because $\pi_i$ can be
viewed as a pure function.

\paragraph{Properties.}
Pairs and the unit value in a lambda calculus typically form a monoid. Assuming $\simeq$ is an
isomorphism that performs appropriate transformation on values, without affecting other
properties (here, coeffects) of the expressions. The monoid laws then correspond to
the demand that $(e_1, (e_2, e_3)) \simeq ((e_1, e_2), e_3)$ (associativity) and the
demand that $((), e) \simeq e \simeq (e, ())$ (unit).

Thanks to the properties of $\cpar$, the flat coeffect calculus obeys the monoid laws for pairs.
In the following, we assume that \ident{assoc} is a pure function transforming a pair $(x_1, (x_2, x_3))$
to a pair $((x_1, x_2), x_3)$. We write $e \equiv e'$ when for all $\Gamma, \tau$ and $\cclrd{r}$,
it is the case that $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ if and only if
$\coctx{\Gamma}{\cclrd{r}} \vdash e' : \tau$.

@@

\noindent
The Remark~\ref{thm:flat-tup-eq} motivates the demand of the monoid structure
$(\C, \cpar, \czero)$ of the flat coeffect algebra. We require only unit and associativity
laws. In our three examples, the $\cpar$ operator is also symmetric, which additionally
gives us the property that $(e_1, e_2) \simeq (e_2, e_1)$.

@@

The flat coeffect calculus is parameterized by a \emph{flat coeffect algebra} that captures
the structure of the information tracked by the type system. We instantiated the system to
capture three specific systems, namely liveness, dataflow and implicit parameters. However,
the system is more general and an capture numerous other applications outlined in
Section~\ref{sec:applications-flat}.

An inherent property of flat coeffect systems is the ambiguity of the typing for lambda
abstraction. The body of a function requires certain context, but the context can be often
provided by either the declaration-site or the call-site. Resolving this ambiguity has to be
done differently for each concrete coeffect system, depending on its specific notion of context.
We discussed this for implicit parameters, dataflow and liveness in Section~\ref{sec}.

Finally, we introduced the equational theory for flat coeffect calculus. Although each
concrete instance of flat coeffect calculus models different notion of context, there are
syntactic properties that hold for all flat coeffect systems satisfying certain additional
conditions. In particular, two \emph{typing preservation} theorems prove that the operational
semantics for two classes of flat coeffect calculi (including liveness and implicit parameters)
can be based on the standard call-by-name reduction.

In the next section, we move from abstract treatment of the flat coeffect calculus to a more
concrete discussion. We explain its category-theoretical motivation, we use it to define
translational semantics (akin to Haskell's do notation) and we prove that well-typed programs
in flat coeffect calcului for implicit parameters and dataflow do not get stuck.

>>>>

\noindent
In two of the three systems, some of the operators of the flat coeffect algebra coincide, but
in the dataflow system all three are distinct. Similarly, the two special elements
coincide in some, but not all systems. The required axioms are motivated by the aim to capture
common properties of the three examples, without unnecessarily restricting the system:

\begin{itemize}
\item The monoid $(\C, \cseq, \cunit)$ represents \emph{sequential} composition of (semantic)
functions. The monoid axioms are required in order to form a category structure in the
semantics (Section~\ref{sec:flat-semantics}).

\item The idempotent monoid $(\C, \cpar, \czero)$ represents \emph{pointwise} composition,
\ie~the case when the same context is passed to multiple (independent) computations. The monoid
axioms guarantee that usual syntactic transformations on tuples and the unit value
(Section~\ref{sec:flat-exts}) preserve the coeffect. Idempotence holds for all our examples
and allows us to unify the flat and structural systems in Chapter~\ref{ch:unified}.

@@

\noindent
For simplicity, assume that all parameters have the same type $\ident{num}$ and so the annotations only
track sets of names. The definition uses a set union for all three operations. Both variables and
constants are are annotated with $\emptyset$ and the ordering is defined by $\subseteq$. The
definition satisfies the flat coeffect algebra axioms because $(S, \cup, \emptyset)$ is an idempotent,
commutative monoid. The language has additional syntax for defining an implicit parameter and for
accessing it, together with associated typing rules:
%
\begin{equation*}
\quad~ e ~::=~ \ldots \sep \ident{?p} \sep \kvd{let}~\ident{?p}~=~e_1~\kvd{in}~e_2
\end{equation*}
\begin{equation*}
\tyrule{param}
  { ~ }
  { \coctx{\Gamma}{\cclrd{ \{ \ident{?p} \} }} \vdash \ident{?p} : \ident{num} }
\end{equation*}
\begin{equation*}
\tyrule{letpar}
  { \coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 &
    \coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_2}
  {\coctx{\Gamma}{\cclrd{r} \;\cclrd{\cup\; (\cclrd{s} \setminus \{ \ident{?p}) \}} } \vdash \kvd{let}~\ident{?p}=e_1~\kvd{in}~e_2 : \tau_2 }
\end{equation*}
%
The (\emph{param}) rule specifies that the accessed parameter $\ident{?p}$ needs to be in the set of
required parameters $\cclrd{r}$. As discussed earlier, we use the same type $\ident{num}$ for all
parameters, but it is also possible to define a coeffect calculus that uses mappings from names to
types (care is needed to avoid assigning multiple types to a parameter of the same type).

@@

As required by the axioms, $(\mathbb{N}, +, 0)$ and $(\mathbb{N}, \mathit{max}, 0)$ form monoids
and $(\mathbb{N}, \mathit{min})$ forms a band. Note that dataflow is our first example where
$\cseq$ is not idempotent. The distributivity axioms require the following to be the case:
$\mathit{max}(r,s) + t = \mathit{max}(r+t, s+t)$, which is easy to see.

@@

\noindent
Note that $(\mathbb{N}_{\bot}, +, 0)$ is a monoid for the extended definition of $+$; for the
bottom element $0 + \bot = \bot$ and for natural numbers $0 + n = n$. The structure
$(\mathbb{N}, \emph{max}, \bot)$ is also a monoid, because $\bot$ is the least element and so
$\mathit{max}(n, \bot) = n$. Finally,  $(\mathbb{N}, \emph{min})$ is a band (the extended
\emph{min} is still idempotent and associative) and the distributivity axioms also hold
for $\mathbb{N}_{\bot}$.

@@

The flat coeffect algebra introduced in Section~\ref{sec:flat-calculus} requires a number of axioms.
The axioms are required for three reasons -- to be able to define the categorical structure
in Section~\ref{sec:semantics-theory}, to prove equational properties in Section~\ref{sec:flat-syntax}
and finally, to guarantee intuitive syntactic properties for constructs such as
$\lambda$-abstraction and pairs in context-aware calculi.

@@

Recall the (\emph{let}) rule in Figure~\ref{fig:flat-types}. It annotates the expression
$\kvd{let}~x=e_1~\kvd{in}~e_2$ with context demands $\cclrd{s}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{r})$.
This rule can be derived from the typing derivation for an expression $(\lambda x.e_2)~e_1$ as
a special case. We use the idempotence of $\czip$ as follows:
%
\begin{equation*}
\tyrule{app}
  {\begin{array}{l}
   \vspace{-1.5em}
   \coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1
   \end{array} &
   \tyruler{abs}
       { \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{s}} \vdash e_2 : \tau_2 }
       { \coctx{\Gamma}{\cclrd{s}} \vdash \lambda x.e_2 : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 } }
  { \coctx{\Gamma}{\cclrd{s}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{r})} \vdash (\lambda x.e_2)~e_1 : \tau_2 }
\end{equation*}
%
This is one possible derivation, but other derivations may be valid for concrete coeffect system.
The design decision of using this particular derivation for the typing of \kvd{let} is motivated by
the fact that the typing obtained using the special rule is more precise for all the examples
consider in this chapter. To see this, assume an arbitrary splitting
$\cclrd{s} = \cclrd{s_1}\,\czip\,\cclrd{s_2}$. Table~\ref{tab:flat-simplelet}
shows the coeffect annotation derived from $(\lambda x.e_2)~e_1$, the coeffect annotation obtained
by the (\emph{let}) rule and the simplified coeffect annotation using the particular flat coeffect
algebras.

@@

It is perhaps somewhat unexpected that the annotation can be simplified in different ways for
different examples. However, for all our systems, the simplfied annotation (right column in
Table~\ref{tab:flat-simplelet}) is more precise than the original (left column). Recall that
$\cclrd{s} = \cclrd{s_1}\,\czip\,\cclrd{s_2}$. The following inequalities hold:
%
\begin{equation*}
\begin{array}{rclll}
\cclrd{s_1} \cup (\cclrd{s_2} \cup \cclrd{r}) &\supseteq& (\cclrd{s_1} \cup \cclrd{s_2}) \cup \cclrd{r}
  && \textnormal{(implicit parameters)}\\
\cclrd{s_1} \sqcap (\cclrd{s_2} \sqcup \cclrd{r}) &\sqsupseteq&  (\cclrd{s_1} \sqcap \cclrd{s_2})
  && \textnormal{(liveness)} \\
\mathit{max}(\cclrd{s_1}, \cclrd{s_2} + \cclrd{r}) &\geq& \mathit{min}(\cclrd{s_1}, \cclrd{s_2}) + \cclrd{r}
  &\quad& \textnormal{(dataflow)} \\
\end{array}
\end{equation*}
%
In other words, the inequality states that using idempotence, we get a more precise typing.
Using the $\cgeq$ operator of flat coeffect algebra, this property can be expressed in general as:
%
\begin{equation*}
\cclrd{s_1} \;\cpar\; (\cclrd{s_2} \,\cseq\, \cclrd{r}) \;\cgeq\;
  (\cclrd{s_1} \,\czip\, \cclrd{s_2}) \;\cpar\; ((\cclrd{s_1} \,\czip\, \cclrd{s_2}) \,\cseq\, \cclrd{r})
\end{equation*}
%
This property does not follow from the axioms of the flat coeffect algebra. To make
the flat coeffect system as general as possible, we do not \emph{in general} require it as
an additional axiom, although the above examples provide a reasonable basis for using
the specialized (\emph{let}) rule in the flat coeffect system.

@@

In Section~\ref{sec:flat-calculus-lambda}, we discussed how to reconcile two typings for
lambda abstraction -- for implicit parameters, the lambda function needs to split context
demands using $\cclrd{r} \cup \cclrd{s}$, but for dataflow and liveness it suffices to
duplicate the demand $\cclrd{r}$ of the body. We consider coeffect calculi for which the
simpler duplication of coeffects is sufficient.

\paragraph{Simplified abstraction.}
Recall that $(\C, \czip)$ is a band, that is, $\czip$  is idempotent and associative. The
idempotence means that the context demands of the body can be required from both the
declaration site and the call site. In Section~\ref{sec:flat-unique-ldf}, we introduced the
(\emph{idabs}) rule (repeated below for reference), which uses the idempotence and duplicates
coeffect annotations:

@@

\noindent
The practical consequence of the Remark~\ref{thm:flat-alt-abs} is that, for strictly
oriented coeffect calculi (such as our liveness and dataflow computations), the (\emph{idabs})
rule not only determines a unique typing derivation (as discussed in Section~\ref{sec:flat-unique-ldf}),
but it gives (together with subtyping and subcoeffecting) an equivalent type system.

@@

\noindent
Intuitively, this means that the context demands of a function with multiple arguments can be
split arbitrarily between the declaration site and (multiple) call sites.

@@

To focus on the key aspects of flat coeffect systems, the calculus introduced in Section~\ref{sec:flat-calculus}
consists only of variables, abstraction, application and let binding. Here, we extend it with pairs
and the unit value to sketch how it can be turned into a more complete programming language and to
further motivate the axioms for $\cpar$. The syntax of the language is extended as follows:

\noindent
\begin{equation*}
\begin{array}{rcl}
e &::=& \ldots \sep () \sep e_1, e_2 \\
\tau &::=& \ldots \sep \ident{unit} \sep \tau \times \tau
\end{array}
\end{equation*}
%
The typing rules for pairs and the unit value are shown in Figure~\ref{fig:flat-ext-types}.
The unit value (\emph{unit}) is annotated with the $\czero$ coeffect (the same as other constants).
Pairs, created using the $(e_1, e_2)$ expression, are annotated with a coeffect that combines
the coeffects of the two sub-expressions using the \emph{pointwise} operator $\cpar$. The operator
models the case when the (same) available context is split and passed to two independent
sub-expressions. Finally, the (\emph{proj}) rule is uninteresting, because $\pi_i$ can be
viewed as a pure unary function.

\paragraph{Properties.}
Pairs and the unit value in a lambda calculus typically form a monoid. Assuming $\simeq$ is an
isomorphism that performs appropriate transformation on values, without affecting other
properties (here, coeffects) of the expressions. The monoid axioms then correspond to
the requirement that $(e_1, (e_2, e_3)) \simeq ((e_1, e_2), e_3)$ (associativity) and the
demand that $((), e) \simeq e \simeq (e, ())$ (unit).

Thanks to the properties of $\cpar$, the flat coeffect calculus obeys the monoid axioms for pairs.
In the following, we assume that \ident{assoc} is a pure function transforming a pair $(x_1, (x_2, x_3))$
to a pair $((x_1, x_2), x_3)$. We write $e \equiv e'$ when for all $\Gamma, \tau$ and $\cclrd{r}$,
it is the case that $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ if and only if
$\coctx{\Gamma}{\cclrd{r}} \vdash e' : \tau$.

@@

\noindent
The Remark~\ref{thm:flat-tup-eq} motivates the demand of the monoid structure
$(\C, \cpar, \czero)$ of the flat coeffect algebra. We require only unit and associativity
axioms. In our three examples, the $\cpar$ operator is also symmetric, which additionally
guarantees that $(e_1, e_2) \simeq (e_2, e_1)$, which is a property that is expected to hold
for $\lambda$-calculus.

@@

The flat coeffect calculus is parameterized by a \emph{flat coeffect algebra} that captures
the structure of the information tracked by the type system. We instantiated the system to
capture three specific systems, namely liveness, dataflow and implicit parameters. However,
the system is more general and an capture various other applications outlined in
Section~\ref{sec:applications-flat}.

An inherent property of flat coeffect systems is the ambiguity of the typing for lambda
abstraction. The body of a function requires certain context, but the context can be often
provided by either the declaration-site or the call-site. Resolving this ambiguity has to be
done differently for each concrete coeffect system, depending on its specific notion of context.
We discussed this for implicit parameters, dataflow and liveness in Section~\ref{sec:flat-unique}
and noted that the result for dataflow and liveness generalizes for any coeffect calculus
with strictly oriented coeffect algebra (Remark~\ref{thm:flat-alt-abs}).

Finally, we introduced the equational theory for flat coeffect calculus. Although each
concrete instance of flat coeffect calculus models different notion of context, there are
syntactic properties that hold for all flat coeffect systems satisfying certain additional
conditions. In particular, two \emph{type preservation} theorems prove that the operational
semantics for two classes of flat coeffect calculi (including liveness and implicit parameters)
can be based on standard call-by-name reduction.

In the next section, we move from abstract treatment of the flat coeffect calculus to a more
concrete discussion. We explain its category-theoretical motivation, we use it to define
translational semantics (akin to Haskell's do notation) and we prove a soundness result that
well-typed programs in flat coeffect calcului for implicit parameters and dataflow do not get
stuck in the translated version.
